---
title: "data_cleaning"
author: "Sharon"
date: "2023-10-05"
output: html_document
---

Library
```{r}
library(tibble)
library(readr)
library(skimr)
library(dplyr)
library(janitor)
```

raw data
```{r}
raw_data <- read_delim("raw_data/anage_data.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE)

data <- clean_names(raw_data)
```


```{r}
raw_data_summary <- raw_data |>
  group_by(Kingdom, Phylum, Class) |>
  summarise()

raw_data_summary
```

Filtering only the data i want to investigate and identify is there any weird data.
```{r}
mammalia <- data[data$class == "Mammalia",]

mammalia_subset <- mammalia[, c("hagrid", "kingdom", "phylum", "class", "order", "family", "genus", "species", "maximum_longevity_yrs", "gestation_incubation_days", "weaning_days", "adult_weight_g")]

mammalia[, c("hagrid", "kingdom", "phylum", "class", "order", "family", "genus", "species", "maximum_longevity_yrs", "gestation_incubation_days", "weaning_days", "adult_weight_g", "metabolic_rate_w")]
```

```{r}
skim(mammalia_subset)
```

```{r}
mammalia_clean <- na.omit(mammalia_subset)

primates_clean <- mammalia_clean[mammalia_clean$order == "Primates",]
rodentia_clean <- mammalia_clean[mammalia_clean$order == "Rodentia",]
```


Find the HAGRID out

Question:
1. Should i filter the data i want here or it is just data cleaning here?
2. confusion of issue in github
3. some graph problem: weaning day --> rodentia (have smaller range) 
4. if i am only investigating correlation of some order in mammal, is it ok for me to say i chose those because they havee larger sample size
5. Some factors were found to be not signifigant in correlation with lifespan, it is againsting some paper